---
title: "National precipitation isotope data processing"
author: "Bruce Dudley"
date: "2025-07-08"
format:
  html:
    self-contained: true       # make standalone HTML
    code-fold: false
    toc: true
    toc-location: left
    toc-title: Table of contents
    number-sections: true
editor: visual
execute: 
  cache: false                 # disable caching to avoid file permission issues
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      comment = "",
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE,
                      dev = 'png', dpi = 300, 
                      fig.height = 5)

library(tidyverse)  # includes purrr for functional programming
library(robustbase)
library(modelr)
library(lubridate)
library(skimr)       # summary of data
library(dotwhisker)  # dot whisker plot for multiple regression results
library(GGally)      # for scatterplots with correlation with ggpairs()
library(leaflet)
library(htmltools)   # for htmlEscape function
library(ggmap)       # get_stamenmap() & ggmap
library(cowplot)     # multiple plots for a figure
library(broom)       # glance(), tidy() for linear regressions
library(scales)
library(ggrepel)     # geom_text_repel()
library(purrr)
# to do map interpolate of sinusoidal parameters
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
# devtools::install_github("ropensci/rnaturalearthhires")
library(rnaturalearthhires)

library(stars)
library(gstat)

library(janitor)  # for get_dupes() function
library(writexl)


library(flextable)
set_flextable_defaults(big.mark = "")

ft <- function(input.data, caption = "", num.rows = Inf) {
  ft <- input.data %>%
    slice_head(n = num.rows) %>%
    flextable() %>%
    autofit() %>%
    theme_zebra() %>%
    set_caption(caption)
  ft
}

source("scripts/auxiliary.code.R")
source("scripts/vcsn.functions.R")

```

# Read in output from previous runs of this code

```{r}
# Setup chunk — load processed climate data
# output_dir <- "output"
# rds_files <- list.files(output_dir, pattern = "^vcsn_data_.*\\.rds$", full.names = TRUE)

# Load into a named list, named by site
# climate_data_all_sites <- lapply(rds_files, readRDS)

# Replace underscores with spaces in names to match isotope Site_Name format
# site_names <- gsub("^vcsn_data_|\\.rds$", "", basename(rds_files))
# site_names_spaces <- gsub("_", " ", site_names)

# names(climate_data_all_sites) <- site_names_spaces


```

# Read in and check first raw dataset

We'll start with the McDonnell dataset and do a few checks on our methods

### Read in data

```{r}

library(readxl)

# Read in the lookup table and sub-daily data
lookup_table <- read_excel("JeffMcD.xlsx", sheet = "SITE DATA lookup")
lookup_table<- lookup_table %>% dplyr::rename(lon = long)
ddata <- read_excel("JeffMcD.xlsx", sheet = "DATA")
dates<-read.csv('dates.csv')
```

### Add lats and longs from lookup to main dataframe

```{r}
library(dplyr)

# Join long and lat from lookup_table into sub_daily_data
ddata <- ddata %>%
  left_join(
    lookup_table %>% select(Site_Name, lon, lat),
    by = "Site_Name"
  )

```

### Amount weighting

```{r}
daily_data <- ddata %>%
  mutate(Date = as.Date(Collection_Date_Time)) %>%
  group_by(Institution, Project, Site_Name, SampleID, Date, Collector_type, lon, lat) %>%
  summarise(
    daily_precip = sum(Precipitation_mm, na.rm = TRUE),
    d2H = {
      valid <- Precipitation_mm > 0 & !is.na(d2H)
      total_weight <- sum(Precipitation_mm[valid], na.rm = TRUE)
      if (total_weight > 0) {
        sum(d2H[valid] * Precipitation_mm[valid], na.rm = TRUE) / total_weight
      } else {
        NA_real_
      }
    },
    d18O = NA_real_,  # Add empty column for compatibility
    .groups = "drop"
  ) %>%
  rename(Precipitation_mm = daily_precip)



```

# Import VCSN data

Now, for checking, we'll bring in the Virtual Climate Station Network (VCSN) data from the nearest grid point. We're most interested in daily rain data, because we'll need that to do rain weighting at sites without rainfall amount measurements.

### Load the VCSN data folder

```{r data VCSN locations}

VCSN.directory <- "C:/Users/dudleybd/OneDrive - NIWA/Desktop/Precip isotope model/vcsn_data/"

vcsn_info(VCSN.directory)
vcsn.agent.locations <- vcsn_agent_locations(VCSN.directory)
vcsn.agent.locations |> slice_head(n = 5)

#plot(vcsn.agent.locations["Agent"], cex = 0.5, pch = 16)

```

The climate fields in the VCSN data are identified as

<https://one.niwa.co.nz/display/CLIDB/VCSN+Times+and+Field+Descriptions>

<https://one.niwa.co.nz/display/SYSTEMS/The+size+of+the+problem>

(a) MSLP (Mean Sea Level Pressure)
(b) PET (Potential Evapotranspiration)
(c) Rain (daily rainfall)
(d) RH (relative humidity)
(e) SoilM (soil moisture)
(f) ETmp (earth temperature at 10cm depth)
(g) Rad (solar radiation)
(h) TMax (maximum temperature)
(i) Tmin (minimum temperature)
(j) VP (vapour pressure)
(k) Wind (average wind speed at 10m above ground level)
(l) Rain_bc (Rainfall with bias correction)
(m) Tmax_N (maximum temperature via different method)
(n) Tmin_N (minimum temperature via different method)

Potential evaporation or potential evapotranspiration is defined as the amount of evaporation that would occur if a sufficient water source were available.

*Rain_bc*, *Tmax_n*, *Tmin_N* were introducted in 2019.

### Append VCSN data to site (point, not catchment) information

First add VCSN Agent numbers to the lookup table for Jeff's data

```{r}


lookup_table.plus.vcsn.agent <- lookup_table |>
  vcsn_append_nearest_vcsn(VCSN.directory) 
slice_head(lookup_table.plus.vcsn.agent, n = 1)
```

Clean up daily_data to match the full data file we'll import below.

```{r}

# Step 1: Rename `Date` to `Collection_Date`
daily_data <- daily_data %>%
  rename(Collection_Date = Date)

# Step 2: Add `Start_Date` as previous collection date for each site
daily_data <- daily_data %>%
  arrange(Site_Name, Collection_Date) %>%
  group_by(Site_Name) %>%
  mutate(Start_Date = lag(Collection_Date)) %>%
  ungroup()


```

### Data extraction for a single site

```{r}

# Step 1: Filter daily_data to this one site (for test)
site_name <- lookup_table.plus.vcsn.agent$Site_Name[1]
daily_data_site <- daily_data %>%
  filter(Site_Name == site_name)

# Step 2: Find earliest Start_Date and latest Collection_Date
date_range <- daily_data_site %>%
  summarise(
    min_date = min(Start_Date, na.rm = TRUE),
    max_date = max(Collection_Date, na.rm = TRUE)
  )

# Step 3: Get the years spanning this range (buffering to full years)
years_needed <- seq(year(date_range$min_date), year(date_range$max_date))

# Step 4: Pull out VCSN data for this site and date range
# Loop over each site in the lookup table
for (this_site in unique(lookup_table$Site_Name)) {

  message("\n=== Starting site: ", this_site, " ===")

  # Subset data
  this_lookup <- lookup_table %>% filter(Site_Name == this_site)
  this_daily <- daily_data %>% filter(Site_Name == this_site)

  # Date range
  date_range <- this_daily %>%
    summarise(
      min_date = min(Start_Date, na.rm = TRUE),
      max_date = max(Collection_Date, na.rm = TRUE)
    )

  years_needed <- seq(year(date_range$min_date), year(date_range$max_date))

  message("Years: ", paste(years_needed, collapse = ", "))

  # Pull VCSN data
  this_climate <- vcsn_append_climate_information(this_lookup, VCSN.directory, years_needed)

  # Filter to actual range
  this_climate <- this_climate %>%
    filter(Date >= date_range$min_date, Date <= date_range$max_date)

  # Save or store
  saveRDS(this_climate, file = paste0("vcsn_data_", gsub(" ", "_", this_site), ".rds"))
  message("✅ Done with site: ", this_site)
}


```

### Function to assign isotope values to daily rainfall

```{r}
assign_isotopes_to_days <- function(site_df, iso_df, site_name) {
  site_climate <- site_df %>% filter(Site_Name == site_name)
  site_isotopes <- iso_df %>% filter(Site_Name == site_name)

  assigned_values <- pmap(
    list(site_isotopes$Start_Date, site_isotopes$Collection_Date, site_isotopes$d2H, site_isotopes$d18O),
    function(start_date, end_date, d2h_val, d18o_val) {
      if (is.na(start_date) | is.na(end_date) | start_date >= end_date) return(NULL)

      tibble(
        Date = seq.Date(start_date, end_date - 1, by = "day"),
        Site_Name = site_name,
        d2H = d2h_val,
        d18O = d18o_val
      )
    }
  )

  assigned_df <- bind_rows(assigned_values)

  left_join(site_climate, assigned_df, by = c("Date", "Site_Name"))
}


```

### Run the function

```{r}
# Set site name (e.g., for your test site)
test_site <- "MAIMAI (M8)"
daily_with_isotopes <- assign_isotopes_to_days(this_climate, daily_data, test_site)

# Check
head(daily_with_isotopes %>% select(Date, Rain_bc, d2H, d18O))

```

### Clip to valid isotope period

We keep only days that fall within the range of the isotope records:

```{r}
# Determine valid isotope date ranges for each site
iso_date_ranges <- daily_data %>%
  group_by(Site_Name) %>%
  summarise(
    min_date = min(Start_Date, na.rm = TRUE),
    max_date = max(Collection_Date, na.rm = TRUE),
    .groups = "drop"
  )

# Clip to these ranges
daily_with_isotopes_clipped <- daily_with_isotopes %>%
  left_join(iso_date_ranges, by = "Site_Name") %>%
  filter(Date >= min_date & Date <= max_date) %>%
  select(-min_date, -max_date)

```

We group by site and month to calculate rainfall-weighted monthly isotope values:

```{r}
monthly_weighted <- daily_with_isotopes_clipped %>%
  mutate(
    year = year(Date),
    month = month(Date),
    year_month = floor_date(Date, "month")
  ) %>%
  group_by(Site_Name, year, month, year_month) %>%
  summarise(
    total_rainfall = sum(Rain_bc, na.rm = TRUE),
weighted_d2H = ifelse(total_rainfall == 0, NA,
                      sum(Rain_bc * d2H, na.rm = TRUE) / total_rainfall),
weighted_d18O = ifelse(total_rainfall == 0, NA,
                       sum(Rain_bc * d18O, na.rm = TRUE) / total_rainfall),
    days_present = n_distinct(Date),
    expected_days = days_in_month(first(Date)),
    missing_days = expected_days - days_present,
    incomplete_month = ifelse(missing_days > 0, "YES", "NO"),
    .groups = "drop"
  )

```

### Plotting back monthly weighted against measured daily data:

```{r plotting back weighted}


# Prepare monthly weighted d2H for plotting (representing *previous month*)
monthly_weighted_d2H_plot <- monthly_weighted %>%
  mutate(
    Date = as.Date(paste(year, month, "01", sep = "-")) %m+% months(1)
  ) %>%
  dplyr::select(Site_Name, Date, d2H = weighted_d2H) %>%
  mutate(Source = "Weighted monthly")

# Prepare daily isotope sample data
precip_iso_d2H_plot <- daily_data %>%
  dplyr::select(Site_Name, Date = Collection_Date, d2H) %>%
  mutate(Source = "Sample measurement")

# Combine
combined_plot_d2H <- bind_rows(monthly_weighted_d2H_plot, precip_iso_d2H_plot) %>%
  filter(!is.na(Date) & !is.na(d2H))

# Plot
p<- ggplot(combined_plot_d2H, aes(x = Date, y = d2H, color = Source)) +
  geom_point(size = 2) +
  geom_line(
    data = filter(combined_plot_d2H, Source == "Weighted monthly"),
    aes(group = Site_Name),
    size = 0.8
  ) +
  facet_wrap(~ Site_Name, scales = "free_x") +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y"
  ) +
  labs(
    x = "Date",
    y = expression(delta^{2}*H~"(\u2030)")
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

ggsave(
  filename = "Maimai weighting effects.tif",
  plot = p,
  device = "tiff",
  width = 10,    # in inches
  height = 6,    # in inches
  dpi = 300
)
p
```

### Revised plot

```{r}


# --- prepare monthly / daily (same as you) ---
monthly_weighted_d2H_plot <- monthly_weighted %>%
  mutate(Date = as.Date(paste(year, month, "01", sep = "-")) %m+% months(1)) %>%
  select(Site_Name, Date, d2H = weighted_d2H, total_rainfall, incomplete_month) %>%
  mutate(Source = "Weighted monthly")

precip_iso_d2H_plot <- daily_data %>%
  select(Site_Name, Date = Collection_Date, d2H, Precipitation_mm) %>%
  mutate(Source = "Sample measurement")

combined_plot_d2H <- bind_rows(monthly_weighted_d2H_plot, precip_iso_d2H_plot) %>%
  filter(!is.na(Date) & !is.na(d2H))

# --- scaling & axis tick locations ---
min_d2H <- min(combined_plot_d2H$d2H, na.rm = TRUE)    # negative value
max_rain <- max(daily_data$Precipitation_mm, na.rm = TRUE)

rain_scale <- (abs(min_d2H) * 0.9) / max_rain

# primary limits and breaks (left axis: negative values up to 0)
primary_min <- floor(min_d2H / 10) * 10
primary_max <- ceiling(max_rain * rain_scale / 10) * 10
primary_breaks <- seq(primary_min, 0, by = 10)

# secondary axis ticks (rain mm)
sec_rain_breaks_mm <- pretty(c(0, max_rain), n = 5)          # e.g. 0,10,20...
sec_primary_breaks <- sec_rain_breaks_mm * rain_scale        # positions in primary coords

# --- x limits: use exact floor/ceiling dates to exclude Feb ---
x_min <- floor_date(min(combined_plot_d2H$Date), "month")
x_max <- ceiling_date(max(monthly_weighted_d2H_plot$Date), "month") - days(1)  # last day of Jan

# --- vertical lines at start of each month within x-axis range ---
vlines <- seq(x_min, x_max, by = "month")

# prepare negative isotope horizontal gridline positions (in primary coords)
iso_grid_ys <- primary_breaks  # e.g. -80, -70, ..., 0

# --- Plot ---

p <- ggplot() +
  # rainfall bars
  geom_col(
    data = daily_data,
    aes(x = Collection_Date, y = Precipitation_mm * rain_scale),
    fill = "steelblue3",
    color = "steelblue4",
    width = 0.8
  ) +
  # isotope points
  geom_point(
    data = filter(combined_plot_d2H, Source == "Sample measurement"),
    aes(x = Date, y = d2H, color = "Sample measurement"),
    size = 2
  ) +
  geom_point(
    data = filter(combined_plot_d2H, Source == "Weighted monthly"),
    aes(x = Date, y = d2H, color = "Weighted monthly"),
    size = 2
  ) +
  geom_line(
    data = filter(combined_plot_d2H, Source == "Weighted monthly"),
    aes(x = Date, y = d2H, color = "Weighted monthly", group = Site_Name),
    linewidth = 0.8
  ) +
  # vertical dashed month lines
  geom_vline(
    xintercept = vlines,
    linetype = "dashed",
    color = "grey70"
  ) +
  # x-axis
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y",
    limits = c(x_min, x_max),
    expand = c(0, 0)
  ) +
  facet_wrap(~ Site_Name, scales = "free_x") +
  # isotope horizontal gridlines
  annotate(
    "segment",
    x = x_min, xend = x_max,
    y = iso_grid_ys,
    yend = iso_grid_ys,
    colour = "grey85",
    size = 0.3
  ) +
  # y-axis scales
  scale_y_continuous(
    name = expression(delta^{2}*H~"(‰)"),
    limits = c(primary_min, primary_max),
    breaks = primary_breaks,
    sec.axis = sec_axis(
      trans = ~ . / rain_scale,
      name = "Precipitation (mm)",
      breaks = sec_primary_breaks,
      labels = sec_rain_breaks_mm
    )
  ) +
  scale_color_manual(
    values = c("Sample measurement" = "grey40", "Weighted monthly" = "#D55E00"),
    name = NULL
  ) +
  labs(x = "Date") +
  theme_minimal(base_size = 12) +
 theme(
  legend.position = "bottom",
  axis.text.x = element_text(angle = 0, hjust = 0.5),
  panel.grid.major.y = element_blank(),
  panel.grid.minor = element_blank(),
  strip.text = element_text(size = 10),
  
  # Primary y-axis (left)
  axis.title.y.left = element_text(
    vjust = -0.5,                 # center vertically
    margin = margin(r = 5)      # push left title away from axis
  ),
  
  # Secondary y-axis (right)
  axis.title.y.right = element_text(
    vjust = 1.5,                 # center vertically
    margin = margin(l = 7)      # push right title away from axis
  ),
  
  plot.title = element_blank(),
  plot.margin = margin(10, 30, 10, 20)
)
# save + show
ggsave("Maimai_weighting_effects_vlines_rainbars_fixed_axes_v2.tif", plot = p,
       device = "tiff", width = 10, height = 6, dpi = 300, compression = "lzw")

p



```

## Another version of the plot, including 3-day and weekly rolling weighted values

```{r}

library(dplyr)
library(slider)
library(lubridate)
library(tidyr)
library(ggplot2)

### ----------------------------------------------------------
### 1. Prepare monthly weighted d2H
### ----------------------------------------------------------
monthly_weighted_d2H_plot <- monthly_weighted %>%
  mutate(Date = as.Date(paste(year, month, "01", sep = "-")) %m+% months(1)) %>%
  select(Site_Name, Date, d2H = weighted_d2H, total_rainfall, incomplete_month) %>%
  mutate(Source = "Weighted monthly")

### ----------------------------------------------------------
### 2. Prepare sample-measured daily isotope values
### ----------------------------------------------------------
precip_iso_d2H_plot <- daily_data %>%
  select(Site_Name, Date = Collection_Date, d2H, Precipitation_mm) %>%
  mutate(Source = "Sample measurement")

### ----------------------------------------------------------
### 3. Compute 3-day and 7-day rolling weighted values
### ----------------------------------------------------------
rolling_weighted <- daily_with_isotopes_clipped %>%
  arrange(Site_Name, Date) %>%
  group_by(Site_Name) %>%
  mutate(
    rain_3d = slide_dbl(
      Rain_bc,
      ~ if (all(is.na(.x))) NA_real_ else sum(.x, na.rm = TRUE),
      .before = 2,
      .complete = TRUE
    ),

    wx_3d = slide_dbl(
      Rain_bc * d2H,
      ~ if (all(is.na(.x))) NA_real_ else sum(.x, na.rm = TRUE),
      .before = 2,
      .complete = TRUE
    ),
    d2H_3d = case_when(
      is.na(rain_3d) ~ NA_real_,
      rain_3d == 0 ~ NA_real_,
      TRUE ~ wx_3d / rain_3d
    ),
    rain_7d = slide_dbl(
      Rain_bc,
      ~ if (all(is.na(.x))) NA_real_ else sum(.x, na.rm = TRUE),
      .before = 6,
      .complete = TRUE
    ),
    wx_7d = slide_dbl(
      Rain_bc * d2H,
      ~ if (all(is.na(.x))) NA_real_ else sum(.x, na.rm = TRUE),
      .before = 6,
      .complete = TRUE
    ),
    d2H_7d = case_when(
      is.na(rain_7d) ~ NA_real_,
      rain_7d == 0 ~ NA_real_,
      TRUE ~ wx_7d / rain_7d
    )
  ) %>%
  ungroup()

### ----------------------------------------------------------
### 4. Combine ALL isotope datasets — ONLY ONCE
### ----------------------------------------------------------

combined_plot_d2H <- bind_rows(
  monthly_weighted_d2H_plot,
  precip_iso_d2H_plot,
  rolling_plot
) %>%
  filter(!is.na(Date) & !is.na(d2H)) %>%
  mutate(
    d2H = ifelse(d2H == 0, NA_real_, d2H)  # replace all exact zeros with NA
  )


### ----------------------------------------------------------
### 5. Axis ranges and scaling
### ----------------------------------------------------------
min_d2H <- min(combined_plot_d2H$d2H, na.rm = TRUE)
max_rain <- max(daily_data$Precipitation_mm, na.rm = TRUE)

rain_scale <- (abs(min_d2H) * 0.9) / max_rain

primary_min <- floor(min_d2H / 10) * 10
primary_max <- ceiling(max_rain * rain_scale / 10) * 10
primary_breaks <- seq(primary_min, 0, by = 10)

sec_rain_breaks_mm <- pretty(c(0, max_rain), n = 5)
sec_primary_breaks  <- sec_rain_breaks_mm * rain_scale

x_min <- floor_date(min(combined_plot_d2H$Date), "month")
x_max <- ceiling_date(max(monthly_weighted_d2H_plot$Date), "month") - days(1)

vlines <- seq(x_min, x_max, by = "month")
iso_grid_ys <- primary_breaks

### ----------------------------------------------------------
### 6. Plot
### ----------------------------------------------------------
p <- ggplot() +
  geom_col(
    data = daily_data,
    aes(x = Collection_Date, y = Precipitation_mm * rain_scale),
    fill = "steelblue3",
    color = "steelblue4",
    width = 0.8
  ) +
  geom_point(
    data = filter(combined_plot_d2H, Source == "Sample measurement"),
    aes(x = Date, y = d2H, color = "Sample measurement"),
    size = 2
  ) +
  geom_point(
    data = filter(combined_plot_d2H, Source == "Weighted monthly"),
    aes(x = Date, y = d2H, color = "Weighted monthly"),
    size = 2
  ) +
  geom_line(
    data = filter(combined_plot_d2H, Source == "Weighted monthly"),
    aes(x = Date, y = d2H, color = "Weighted monthly", group = Site_Name),
    linewidth = 0.8
  ) +
  geom_line(
    data = filter(combined_plot_d2H, Source == "Weighted 3-day"),
    aes(x = Date, y = d2H, color = "Weighted 3-day", group = Site_Name),
    linewidth = 0.6, alpha = 0.8
  ) +
  geom_line(
    data = filter(combined_plot_d2H, Source == "Weighted 7-day"),
    aes(x = Date, y = d2H, color = "Weighted 7-day", group = Site_Name),
    linewidth = 0.6, alpha = 0.8
  ) +
  geom_vline(xintercept = vlines, linetype = "dashed", color = "grey70") +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y",
    limits = c(x_min, x_max),
    expand = c(0, 0)
  ) +
  facet_wrap(~ Site_Name, scales = "free_x") +
  annotate("segment",
           x = x_min, xend = x_max,
           y = iso_grid_ys, yend = iso_grid_ys,
           colour = "grey85", size = 0.3) +
  scale_y_continuous(
    name = expression(delta^{2}*H~"(‰)"),
    limits = c(primary_min, primary_max),
    breaks = primary_breaks,
    sec.axis = sec_axis(
      trans = ~ . / rain_scale,
      name = "Precipitation (mm)",
      breaks = sec_primary_breaks,
      labels = sec_rain_breaks_mm
    )
  ) +
  scale_color_manual(
    values = c(
      "Sample measurement" = "grey40",
      "Weighted monthly"   = "#D55E00",
      "Weighted 3-day"     = "#0072B2",
      "Weighted 7-day"     = "#009E73"
    ),
    name = NULL
  ) +
  labs(x = "Date") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 10),
    axis.title.y.left = element_text(vjust = -0.5, margin = margin(r = 5)),
    axis.title.y.right = element_text(vjust = 1.5, margin = margin(l = 7)),
    plot.margin = margin(10, 30, 10, 20)
  )

ggsave("Maimai_weighting_effects_vlines_rainbars_fixed_axes_v2.tif",
       plot = p, device = "tiff", width = 10, height = 6, dpi = 300, compression = "lzw")

p

```

## Make a gif!

Step 1: Install / load necessary packages

```{r}
library(gganimate)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

# --- 1. Prepare rainfall as a source ---
rain_anim <- daily_data %>%
  select(Site_Name, frame_date = Collection_Date, Precipitation_mm) %>%
  mutate(Source = "Rain") %>%
  mutate(d2H = NA_real_)  # add d2H column for compatibility

# --- 2. Prepare sample measurements ---
sample_anim <- combined_plot_d2H %>%
  filter(Source == "Sample measurement") %>%
  select(Site_Name, frame_date = Date, d2H, Source, Precipitation_mm)  # keep Precip_mm for later

# --- 3. Prepare monthly & rolling weights ---
weighted_anim <- combined_plot_d2H %>%
  filter(Source != "Sample measurement" & Source != "Rain") %>%
  select(Site_Name, frame_date = Date, d2H, Source) %>%
  mutate(Precipitation_mm = NA_real_)

# --- 4. Combine everything ---
combined_anim <- bind_rows(rain_anim, sample_anim, weighted_anim) %>%
  arrange(Site_Name, frame_date)

# --- 5. Axis scaling ---
min_d2H <- min(combined_anim$d2H, na.rm = TRUE)
max_rain <- max(daily_data$Precipitation_mm, na.rm = TRUE)
rain_scale <- (abs(min_d2H) * 0.9) / max_rain

primary_min <- floor(min_d2H / 10) * 10
primary_max <- ceiling(max_rain * rain_scale / 10) * 10
primary_breaks <- seq(primary_min, 0, by = 10)

sec_rain_breaks_mm <- pretty(c(0, max_rain), n = 5)
sec_primary_breaks  <- sec_rain_breaks_mm * rain_scale

x_min <- min(combined_anim$frame_date)
x_max <- max(combined_anim$frame_date)
vlines <- seq(x_min, x_max, by = "month")
iso_grid_ys <- primary_breaks

# --- 6. Animated plot ---
p_anim <- ggplot(combined_anim) +
  # Rainfall bars
  geom_col(
    data = filter(combined_anim, Source == "Rain"),
    aes(x = frame_date, y = Precipitation_mm * rain_scale),
    fill = "steelblue3",
    color = "steelblue4",
    width = 0.8
  ) +
  # Points for daily & monthly
  geom_point(
    data = filter(combined_anim, Source != "Rain"),
    aes(x = frame_date, y = d2H, color = Source),
    size = 2
  ) +
  # Lines for weighted
  geom_line(
    data = filter(combined_anim, Source %in% c("Weighted monthly", "Weighted 3-day", "Weighted 7-day")),
    aes(x = frame_date, y = d2H, color = Source, group = Site_Name),
    linewidth = 0.8
  ) +
  geom_vline(xintercept = vlines, linetype = "dashed", color = "grey70") +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y",
    limits = c(x_min, x_max)
  ) +
  scale_y_continuous(
    name = expression(delta^{2}*H~"(‰)"),
    limits = c(primary_min, primary_max),
    breaks = primary_breaks,
    sec.axis = sec_axis(~ ./rain_scale, name = "Precipitation (mm)", breaks = sec_primary_breaks, labels = sec_rain_breaks_mm)
  ) +
  scale_color_manual(
    values = c(
      "Sample measurement" = "grey40",
      "Weighted monthly"   = "#D55E00",
      "Weighted 3-day"     = "#0072B2",
      "Weighted 7-day"     = "#009E73"
    ),
    name = NULL
  ) +
  facet_wrap(~ Site_Name, scales = "free_x") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 10),
    axis.title.y.left = element_text(vjust = -0.5, margin = margin(r = 5)),
    axis.title.y.right = element_text(vjust = 1.5, margin = margin(l = 7))
  ) +
  # animate over date
  transition_reveal(frame_date) +
  shadow_mark(past = TRUE, future = FALSE, alpha = 0.5)

# --- 7. Save animation ---
anim_save("Maimai_d2H_animation.gif", p_anim, width = 1000, height = 600, fps = 10)


```

# Import complete database

```{r}
 library(readxl)
Lookup_full_database <- read.csv("Full_lookup.csv")
# View(Lookup_full_database)
 
 Full_database <- read.csv("Full_dailyplus.csv")
# View(Full_database)
 
```

# Database and lookup table cleaning

### Check which Site_Name values in Full_database don't match any in the lookup table

```{r}
# Get unmatched site names from Full_database
bad_sites_in_data <- setdiff(
  unique(Full_database$Site_Name),
  unique(Lookup_full_database$Site_Name)
)

# View them
bad_sites_in_data
```

Check for any instances of multiple rows with the same Site_Name and Collection_Date

```{r}

# Initial check, after which I went to the original dataframe and deleted 10 rows


#duplicates <- Full_database %>%
 # group_by(Site_Name, Collection_Date) %>%
 # filter(n() > 1) %>%
 # arrange(Site_Name, Collection_Date)

#duplicates

#duplicate_starts <- Full_database %>%
 # group_by(Site_Name, Start_Date) %>%
 # filter(n() > 1) %>%
 #arrange(Site_Name, Start_Date)

#duplicate_starts


any(duplicated(Full_database[, c("Site_Name", "Collection_Date")]))
any(duplicated(Full_database[, c("Site_Name", "Start_Date")]))
```

Check which Site_Name values in the lookup table don't appear in Full_database

```{r}
# Get unmatched site names from Lookup table
unused_lookup_sites <- setdiff(
  unique(Lookup_full_database$Site_Name),
  unique(Full_database$Site_Name)
)

Lookup_full_database %>%
  filter(Site_Name %in% unused_lookup_sites) %>%
  select(Site_Name, Project) %>%
  distinct()
```

The sites with no data are either duplicates (Christchurch and Kaitoke GNIP data are in the Taylor 1990 dataset), snow sites where we can't work out the date the snow fell, and one from Tim Kerr's postdoc that didn't have appropriate data. We can delete all these from the lookup table.

### Clean out unneeded site lookup rows

```{r}
Lookup_filtered <- Lookup_full_database %>%
  filter(!(Site_Name %in% unused_lookup_sites))
```

### Add in missing NZTM and NZMG to lookup table

```{r}
library(sf)
lookup_sf <- st_as_sf(Lookup_filtered, coords = c("long", "lat"), crs = 4326)
# Transform to NZTM
lookup_nztm <- st_transform(lookup_sf, crs = 2193)

# Extract NZTM coordinates
coords_nztm <- st_coordinates(lookup_nztm)

# Add to original dataframe
Lookup_filtered$NZTME <- coords_nztm[,1]  # Easting
Lookup_filtered$NZTMN <- coords_nztm[,2]  # Northing

# Transform to NZMG (legacy)
lookup_nzmg <- st_transform(lookup_sf, crs = 27200)

# Extract and add NZMG coords
coords_nzmg <- st_coordinates(lookup_nzmg)
Lookup_filtered$NZMGE <- coords_nzmg[,1]
Lookup_filtered$NZMGN <- coords_nzmg[,2]


```

### Make reasonable project names for mapping

```{r}
project_name_map <- c(
  "CDRP" = "CDRP",
  "None" = "TUD",
  "GNIP/M/NZ/01" = "GNIP",
  "Forest Flows" = "Forest Flows",
  "Mt TeAroha " = "Mt Te Aroha",
  "Waihi Estuary Focus Catchment Investigation" = "Waihi Estuary",
  "Land Use Intensi?cation: Sustainable Management of Water Quantity and Quality Programme (NIWA)" = "LUWQ NIWA",
  "Rotorua Lakes FaCTS" = "Rotorua FaCTS",
  "NZWaM" = "NZWaM",
  "Taylor 1990" = "DSIR",
  "Tritium Input Function" = "GNS Tritium Input Function",
  "Phase1_early M8 catchment-scale research and observations (1974\x961988)" = "Maimai M8",
  "Hill AP transect" = "Hill AP transect",
  "Tim Kerr postdoc" = "Kerr Postdoc",
  "Hill AP transect / Jacks Thesis" = "Hill AP transect",
  "Waterscape" = "Waterscape",
  "Jack thesis + Zane thesis" = "UC Postgrad",
  "AOSnow" = "Hill AOSnow",
  "Jack thesis" = "UC Postgrad",
  "Ashley Dene" = "Ashley Dene",
  "Hill Haast transect" = "Hill Haast transect",
  "Environment Southland" = "Environment Southland",
  "B3" = "B3",
  "Nehemy/Simmons tree water sources " = "Trent University"
)

# Apply the updated project name map
Lookup_filtered$Project_clean <- dplyr::recode(Lookup_filtered$Project, !!!project_name_map)


```

## Site map showing sites split by project

```{r}
source('scripts/auxiliary.code_BD.R')
ggmap::register_stadiamaps(key = "3ee03876-1ca3-4727-9391-cb8475022ba2")

Lookup_filtered$Color <- Lookup_filtered$Project_clean

# Add a numbered site label from north to south
Lookup_filtered <- Lookup_filtered %>%
  arrange(desc(lat)) %>%
  mutate(Site_Label = row_number())

site_map <- iso_plot_site_locations_BD(
  the.data = Lookup_filtered,
  site.column.name = "Site_Label",
  lat.column.name = "lat",
  lon.column.name = "long",
  color.column.name = "Color",
  island = "NZ"
)

print(site_map)

ggsave("outputs/site_map.tif", site_map, width = 10, height = 8)
```

Looking good. Next step is to....

# Load VCSN climate data for isotope data date ranges for all sites

### Join long/lat to main dataset

We’ll add lat and long columns to Full_database using the Site_Name as the join key:

```{r}

Full_database <- Full_database %>%
  left_join(
    Lookup_filtered %>% select(Site_Name, long, lat),
    by = "Site_Name"
  )
  
```

### Add VCSN Agent (nearest climate grid point) to each site

This attaches the nearest VCSN grid point to each site in the Lookup_filtered table:

```{r}
lookup_with_vcsn <- Lookup_filtered %>%
  rename(lon = long) %>%
  vcsn_append_nearest_vcsn(VCSN.directory)

```

### Check date formats on start and collection dates

```{r}

# Step 1: Check current column classes
str(Full_database[, c("Start_Date", "Collection_Date")])

```

Convert date strings to Date format

```{r}

Full_database <- Full_database %>%
  mutate(
    Start_Date = as.Date(Start_Date, format = "%d/%m/%Y"),
    Collection_Date = as.Date(Collection_Date, format = "%d/%m/%Y")
  )

```

Check for NAs and remove the rows with them...

```{r}
Full_database %>%
  filter(is.na(Start_Date) | is.na(Collection_Date)) %>%
  select(Site_Name, SampleID, Start_Date, Collection_Date)

Full_database <- Full_database %>%
  filter(!is.na(Start_Date) & !is.na(Collection_Date))


```

Check for American interference in date formats!!

This step checks for each site, whether the Start_Date or Collection_Date ever go backward in time, which would strongly suggest a date parsing issue.

```{r}
# Ensure dates are only converted from character to Date
Full_database <- Full_database %>%
  mutate(
    Start_Date = if (is.character(Start_Date)) dmy(Start_Date) else Start_Date,
    Collection_Date = if (is.character(Collection_Date)) dmy(Collection_Date) else Collection_Date
  )

# Find sites with non-sequential Collection_Date
bad_collection <- Full_database %>%
  group_by(Site_Name) %>%
  arrange(Site_Name, Collection_Date) %>%
  mutate(date_diff = Collection_Date - lag(Collection_Date)) %>%
  filter(!is.na(date_diff) & date_diff < 0) %>%
  ungroup()

# Find sites with non-sequential Start_Date
bad_start <- Full_database %>%
  group_by(Site_Name) %>%
  arrange(Site_Name, Start_Date) %>%
  mutate(date_diff = Start_Date - lag(Start_Date)) %>%
  filter(!is.na(date_diff) & date_diff < 0) %>%
  ungroup()

# Print a summary
cat("🚨 Sites with Collection_Date going backward:\n")
print(unique(bad_collection$Site_Name))

cat("\n🚨 Sites with Start_Date going backward:\n")
print(unique(bad_start$Site_Name))


```

## The actual VCSN climate data extraction

### First test for a single site

```{r}
# Select the first unique site name
#this_site <- unique(lookup_with_vcsn$Site_Name)[1]
this_site <- 'Camp Stream'
# Print the site name explicitly
cat("\n=== Starting site:", this_site, "===\n")

# Subset lookup and isotope data for this site
this_lookup <- lookup_with_vcsn %>% filter(Site_Name == this_site)
this_isotope <- Full_database %>% filter(Site_Name == this_site)

# Sanity check
cat("  → Isotope rows:", nrow(this_isotope), "\n")
cat("  → Start_Date NA count:", sum(is.na(this_isotope$Start_Date)), "/", nrow(this_isotope), "\n")

if (nrow(this_isotope) == 0 || all(is.na(this_isotope$Start_Date))) {
  cat("⚠️  No data or missing Start_Date — skipping site.\n")
} else {

  # Determine date range of measurements
  date_range <- this_isotope %>%
    summarise(
      min_date = min(Start_Date, na.rm = TRUE),
      max_date = max(Collection_Date, na.rm = TRUE)
    )

  years_needed <- seq(year(date_range$min_date), year(date_range$max_date))
  cat("  → Years needed:", paste(years_needed, collapse = ", "), "\n")

  # Pull VCSN data
  this_climate <- vcsn_append_climate_information(this_lookup, VCSN.directory, years_needed)

  # Filter to the actual collection period
  this_climate <- this_climate %>%
    filter(Date >= date_range$min_date, Date <= date_range$max_date)

  # Show head of result
  cat("\n✅ Done with site:", this_site, "\n\n")
  print(head(this_climate))
}

```

### Full extraction

```{r}

# Create output directory if it doesn't exist
output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Loop over each site in the lookup table
for (this_site in unique(lookup_with_vcsn$Site_Name)) {

  # Generate expected filename
  outfile <- file.path(output_dir, paste0("vcsn_data_", gsub("[^A-Za-z0-9_]", "", gsub(" ", "_", this_site)), ".rds"))

  # Skip if file already exists
  if (file.exists(outfile)) {
    cat("\n⏭️ Skipping site (already processed):", this_site, "\n")
    next
  }

  start_time <- Sys.time()
  cat("\n=== Starting site:", this_site, "===\n")

  # Subset lookup and isotope data for this site
  this_lookup <- lookup_with_vcsn %>% filter(Site_Name == this_site)
  this_isotope <- Full_database %>% filter(Site_Name == this_site)

  cat("  → Isotope rows:", nrow(this_isotope), "\n")
  cat("  → Start_Date NA count:", sum(is.na(this_isotope$Start_Date)), "/", nrow(this_isotope), "\n")

  if (nrow(this_isotope) == 0 || all(is.na(this_isotope$Start_Date))) {
    cat("⚠️  No data or missing Start_Date — skipping site.\n")
    next
  }

  # Determine date range of measurements
  date_range <- this_isotope %>%
    summarise(
      min_date = min(Start_Date, na.rm = TRUE),
      max_date = max(Collection_Date, na.rm = TRUE)
    )

  years_needed <- seq(year(date_range$min_date), year(date_range$max_date))
  cat("  → Years needed:", paste(years_needed, collapse = ", "), "\n")

  # Pull VCSN data
  this_climate <- vcsn_append_climate_information(this_lookup, VCSN.directory, years_needed)

  # Filter to the actual collection period
  this_climate <- this_climate %>%
    filter(Date >= date_range$min_date, Date <= date_range$max_date)

  # Save the data
  saveRDS(this_climate, file = outfile)

  duration <- round(difftime(Sys.time(), start_time, units = "secs"), 2)
  cat("✅ Done with site:", this_site, "| Duration:", duration, "seconds\n")
}

```

# Stack all the data together

```{r}

# Combine all dataframes in the list into one, adding site name as a column
climate_data_all <- map2_dfr(
  .x = climate_data_all_sites,
  .y = names(climate_data_all_sites),
  .f = ~ mutate(.x, Site_Label_From_ListName = .y)
)
# Tidy up

climate_data_all <- climate_data_all %>%
  select(-`VCSN.lat...24`, -`VCSN.lon...25`) %>%  # remove duplicate columns
  rename(
    VCSN.lat = `VCSN.lat...22`,
    VCSN.lon = `VCSN.lon...23`
  )

head(climate_data_all)
```

# Revised function to assign isotope values to each rainfall day within a collection period

```{r}
# Make sure dates are Date class
Full_database <- Full_database %>%
  mutate(
    Start_Date = as.Date(Start_Date),
    Collection_Date = as.Date(Collection_Date)
  )

# Function stays the same — no changes needed
assign_isotopes_to_days <- function(site_df, iso_df, site_name) {
  site_climate <- site_df %>% filter(Site_Name == site_name)
  site_isotopes <- iso_df %>% filter(Site_Name == site_name)

  assigned_values <- pmap(
    list(site_isotopes$Start_Date, site_isotopes$Collection_Date, site_isotopes$d2H, site_isotopes$d18O),
    function(start_date, end_date, d2h_val, d18o_val) {
      if (is.na(start_date) | is.na(end_date) | start_date >= end_date) return(NULL)

      tibble(
        Date = seq.Date(start_date, end_date - 1, by = "day"),
        Site_Name = site_name,
        d2H = d2h_val,
        d18O = d18o_val
      )
    }
  )

  assigned_df <- bind_rows(assigned_values)

  left_join(site_climate, assigned_df, by = c("Date", "Site_Name"))
}

# Apply to the first site
first_site <- "Camp Stream"

assigned_data <- assign_isotopes_to_days(climate_data_all, Full_database, first_site)

```

Clean out one problematic site

```{r}
# Remove the site from both datasets
Full_database <- Full_database %>% filter(Site_Name != "Pokopoko at Allport Rd")
climate_data_all <- climate_data_all %>% filter(Site_Name != "Pokopoko at Allport Rd")

```

# Full loop to assign isotope values to daily precipitation amounts

```{r}
# Make sure date columns are Date class
Full_database <- Full_database %>%
  mutate(
    Start_Date = as.Date(Start_Date),
    Collection_Date = as.Date(Collection_Date)
  )

assign_isotopes_to_days <- function(site_df, iso_df, site_name) {
  site_climate <- site_df %>% filter(Site_Name == site_name)
  site_isotopes <- iso_df %>% filter(Site_Name == site_name)

  assigned_values <- pmap(
    list(site_isotopes$Start_Date, site_isotopes$Collection_Date, site_isotopes$d2H, site_isotopes$d18O),
    function(start_date, end_date, d2h_val, d18o_val) {
      if (is.na(start_date) | is.na(end_date)) return(NULL)

      date_seq <- if (start_date < end_date) {
        seq.Date(start_date, end_date - 1, by = "day")  # normal case
      } else if (start_date == end_date) {
        start_date  # single-day case
      } else {
        return(NULL)  # invalid range
      }

      tibble(
        Date = date_seq,
        Site_Name = site_name,
        d2H = d2h_val,
        d18O = d18o_val
      )
    }
  )

  assigned_df <- bind_rows(assigned_values)

  if (nrow(assigned_df) == 0) {
    message("⚠️ No valid date range for site: ", site_name)
    return(site_climate %>% mutate(d2H = NA_real_, d18O = NA_real_))
  } else {
    return(left_join(site_climate, assigned_df, by = c("Date", "Site_Name")))
  }
}


# Loop over all unique sites and bind results
all_sites <- unique(Full_database$Site_Name)

assigned_all_sites <- map_dfr(all_sites, function(site) {
  cat("Processing site:", site, "\n")
  assign_isotopes_to_days(climate_data_all, Full_database, site)
})

```

# Generate monthly rainfall weighted values

```{r}
monthly_weighted <- assigned_all_sites %>%
  mutate(
    Rain_bc = as.numeric(Rain_bc),
    d2H = as.numeric(d2H),
    d18O = as.numeric(d18O),
    year = year(Date),
    month = month(Date),
    year_month = floor_date(Date, "month")
  ) %>%
  # Keep all data for plotting completeness
  group_by(Site_Name, year, month, year_month) %>%
  summarise(
    total_rainfall = sum(Rain_bc, na.rm = TRUE),
    weighted_d2H = if (all(is.na(d2H)) || sum(Rain_bc, na.rm = TRUE) == 0) NA else
      sum(Rain_bc * d2H, na.rm = TRUE) / sum(Rain_bc, na.rm = TRUE),
    weighted_d18O = if (all(is.na(d18O)) || sum(Rain_bc, na.rm = TRUE) == 0) NA else
      sum(Rain_bc * d18O, na.rm = TRUE) / sum(Rain_bc, na.rm = TRUE),
    
    days_present = n_distinct(Date),
    expected_days = days_in_month(first(Date)),
    missing_days = expected_days - days_present,
    incomplete_month = ifelse(missing_days > 0, "YES", "NO"),
    sparse_month = days_present < 20,
    .groups = "drop"
  )


```

# Plotting back against measured values

```{r}

# 0. Remove sparse monthly values
monthly_weighted_filtered <- monthly_weighted %>%
  filter(!sparse_month) %>%
  group_by(Site_Name) %>%
  filter(n() > 0) %>%  # Only keep sites with valid data
  ungroup()

# ==== d2H ====

# 1. Monthly weighted d2H for plotting: set date to 1st of next month
monthly_weighted_d2H_plot <- monthly_weighted_filtered %>%
  mutate(Date = make_date(year, month, 1) %>% ceiling_date(unit = "month")) %>%
  select(Site_Name, Date, d2H = weighted_d2H) %>%
  mutate(Source = "Weighted monthly")

# 2. Sample d2H measurements
precip_iso_d2H_plot <- Full_database %>%
  mutate(Collection_Date = as.Date(Collection_Date)) %>%
  select(Site_Name, Date = Collection_Date, d2H) %>%
  mutate(d2H = as.numeric(d2H), Source = "Sample measurement")

# 3. Combine
valid_sites_d2H <- unique(monthly_weighted_d2H_plot$Site_Name)
combined_plot_d2H <- bind_rows(
  monthly_weighted_d2H_plot,
  precip_iso_d2H_plot %>% filter(Site_Name %in% valid_sites_d2H)
) %>%
  filter(!is.na(Date) & !is.na(d2H))

# 4. Save d2H plots
dir.create("plots", showWarnings = FALSE)
pdf("plots/d2H_weighted_vs_measured_by_site.pdf", width = 10, height = 6)

for (site in unique(combined_plot_d2H$Site_Name)) {
  this_data <- combined_plot_d2H %>% filter(Site_Name == site)
  
  p <- ggplot(this_data, aes(x = Date, y = d2H, color = Source, shape = Source)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_line(data = filter(this_data, Source == "Weighted monthly"), aes(group = 1), linewidth = 1) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(
      title = paste0("Measured vs Weighted δ²H - ", site),
      x = "Date",
      y = expression(delta^{2}*H~"(\u2030)"),
      color = "Data Source",
      shape = "Data Source"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 10)
    )
  
  print(p)
}
dev.off()

# ==== d18O ====

# 5. Monthly weighted d18O
monthly_weighted_d18O_plot <- monthly_weighted_filtered %>%
  mutate(Date = make_date(year, month, 1) %>% ceiling_date(unit = "month")) %>%
  select(Site_Name, Date, d18O = weighted_d18O) %>%
  mutate(Source = "Weighted monthly")

# 6. Sample d18O measurements
precip_iso_d18O_plot <- Full_database %>%
  mutate(Collection_Date = as.Date(Collection_Date)) %>%
  select(Site_Name, Date = Collection_Date, d18O) %>%
  mutate(d18O = as.numeric(d18O), Source = "Sample measurement")

# 7. Combine
valid_sites_d18O <- unique(monthly_weighted_d18O_plot$Site_Name)
combined_plot_d18O <- bind_rows(
  monthly_weighted_d18O_plot,
  precip_iso_d18O_plot %>% filter(Site_Name %in% valid_sites_d18O)
) %>%
  filter(!is.na(Date) & !is.na(d18O))

# 8. Save d18O plots
pdf("plots/d18O_weighted_vs_measured_by_site.pdf", width = 10, height = 6)

for (site in unique(combined_plot_d18O$Site_Name)) {
  this_data <- combined_plot_d18O %>% filter(Site_Name == site)
  
  p <- ggplot(this_data, aes(x = Date, y = d18O, color = Source, shape = Source)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_line(data = filter(this_data, Source == "Weighted monthly"), aes(group = 1), linewidth = 1) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(
      title = paste0("Measured vs Weighted δ¹⁸O - ", site),
      x = "Date",
      y = expression(delta^{18}*O~"(\u2030)"),
      color = "Data Source",
      shape = "Data Source"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 10)
    )
  
  print(p)
}
dev.off()


```

# Database add ins

Now we create the two new logical columns in monthly_weighted_filtered dataframe — composite and exact_month — we can use Full_database to check for: 1. composite == TRUE if a month’s data is the result of multiple samples (i.e. multiple rows in Full_database) covering that month at a site. 2. exact_month == TRUE if the data covers the full calendar month exactly:

```         
The earliest Start_Date in that month is the first day of the month, and

The latest Collection_Date is the last day of the month or the first day of the following month.
```

```{r}
# 1. Prepare Full_database with month info
full_with_months <- Full_database %>%
  mutate(
    Start_Date = as.Date(Start_Date),
    Collection_Date = as.Date(Collection_Date),
    year = year(Start_Date),
    month = month(Start_Date)
  )

# 2. Summarise to get composite and exact_month flags
monthly_flags <- full_with_months %>%
  group_by(Site_Name, year, month) %>%
  summarise(
    n_samples = n(),
    min_start = min(Start_Date, na.rm = TRUE),
    max_end = max(Collection_Date, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    composite = n_samples > 1,
    month_start = as.Date(sprintf("%d-%02d-01", year, month)),
    month_end = ceiling_date(month_start, "month") - days(1),
    next_month_start = ceiling_date(month_start, "month"),
    
    exact_month = min_start == month_start & 
      (max_end == month_end | max_end == next_month_start)
  ) %>%
  select(Site_Name, year, month, composite, exact_month)

# 3. Join flags back to your monthly_weighted_filtered dataframe
monthly_weighted_flagged <- monthly_weighted_filtered %>%
  left_join(monthly_flags, by = c("Site_Name", "year", "month")) %>%
  mutate(
    composite = case_when(
      is.na(composite) ~ FALSE,                           # default
      TRUE ~ composite                                     # use original composite flag if available
    ),
    exact_month = ifelse(is.na(exact_month), FALSE, exact_month),
    
    # second-tier: classify as composite if not exact_month but still complete
    composite = ifelse(!exact_month & incomplete_month == "NO", TRUE, composite)
  )

head(monthly_weighted_flagged)
```

Check this is working:

```{r}
#monthly_weighted_flagged %>%
#  filter(Site_Name %in% c("MAIMAI (M8)", "Kaitoke Rain"))

#monthly_weighted_flagged %>%
 # filter(Site_Name == "MAIMAI (M8)")

#assigned_all_sites %>%
#  filter(Site_Name == "MAIMAI (M8)")



```

### Table of different data types in the final weighted dataset

```{r}
# Step 1: Join project information into the flagged monthly data
monthly_with_project <- monthly_weighted_flagged %>%
  left_join(Lookup_filtered %>% select(Site_Name, Project_clean), by = "Site_Name")

# Step 2: Summarise by project
summary_by_project <- monthly_with_project %>%
  group_by(Project_clean) %>%
  summarise(
    n_sites = n_distinct(Site_Name),
    complete_months = sum(incomplete_month == "NO", na.rm = TRUE),
    non_sparse_months = sum(sparse_month == FALSE, na.rm = TRUE),
    exact_months = sum(exact_month == TRUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_sites))  # Optional: sort by number of sites
summary_by_project
```

# Testing for the effects of rainfall weighting using gridded climate data

1.  Calculate a 'true rainfall weighting' column for monthly_weighted_flagged

First we need to clear the 'measured' values out from the CDRP sites - these are actually VCSN data.

```{r}

Full_database <- Full_database %>%
  mutate(Precipitation_mm = as.numeric(Precipitation_mm)) %>%
  mutate(Precipitation_mm = if_else(Project == "CDRP", NA_real_, Precipitation_mm))


```

Then create a dataframe with quasi-daily precipitation isotope data and precipitaion amount data

```{r}
selected_sites <- c("B3. Kaitaia", "B3.Christchurch", "Ashley Dene"
                                     ,"MAIMAI (M8)", "Melrose Farm")

site_rain_data <- Full_database %>%
  filter(Site_Name %in% selected_sites)

```

Add in some precip. data for Ashley Dene, and two of the B3 sites

```{r}
deneprecip<-read.csv("precip/Ashley Dene precipitation.csv")
kaitaiaprecip<-read.csv("precip/kaitaia.csv")
niwachch<-read.csv("precip/niwachch.csv")


# Ensure Date columns are in Date format
site_rain_data <- site_rain_data %>%
  mutate(Start_Date = as.Date(Start_Date),
         Collection_Date = as.Date(Collection_Date))

deneprecip <- deneprecip %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))

kataiaprecip <- kaitaiaprecip %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))

niwachch <- niwachch %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))

# --- Ashley Dene
ashley_dene_samples <- site_rain_data %>%
  filter(Project == "Ashley Dene") %>%
  rowwise() %>%
  mutate(precip_sum = sum(
    deneprecip$Precipitation_mm[
      deneprecip$Date >= Start_Date & deneprecip$Date < Collection_Date
    ],
    na.rm = TRUE
  )) %>%
  ungroup()

# --- B3. Kaitaia
kaitaia_samples <- site_rain_data %>%
  filter(Site_Name == "B3. Kaitaia") %>%
  rowwise() %>%
  mutate(precip_sum = sum(
    kataiaprecip$Rainfall.mm[
      kataiaprecip$Date >= Start_Date & kataiaprecip$Date < Collection_Date
    ],
    na.rm = TRUE
  )) %>%
  ungroup()

# --- B3.Christchurch
chch_samples <- site_rain_data %>%
  filter(Site_Name == "B3.Christchurch") %>%
  rowwise() %>%
  mutate(precip_sum = sum(
    niwachch$Rain.mm[
      niwachch$Date >= Start_Date & niwachch$Date < Collection_Date
    ],
    na.rm = TRUE
  )) %>%
  ungroup()

# Combine and update Precipitation_mm
updated_precip <- bind_rows(
  ashley_dene_samples %>% select(Site_Name, SampleID, precip_sum),
  kaitaia_samples %>% select(Site_Name, SampleID, precip_sum),
  chch_samples %>% select(Site_Name, SampleID, precip_sum)
)

site_rain_data <- site_rain_data %>%
  mutate(Precipitation_mm = as.numeric(Precipitation_mm)) %>%
  left_join(updated_precip, by = c("Site_Name", "SampleID")) %>%
  mutate(Precipitation_mm = if_else(!is.na(precip_sum), precip_sum, Precipitation_mm)) %>%
  select(-precip_sum)


```

Check for rows with missing precipitation data, and count total rows and which sites are covered

```{r}
# Count total missing values
site_rain_data %>%
  filter(is.na(as.numeric(Precipitation_mm))) %>%
  summarise(n_missing = n())

# Count missing values by site
site_rain_data %>%
  filter(is.na(as.numeric(Precipitation_mm))) %>%
  count(Site_Name, sort = TRUE)

nrow(site_rain_data)
unique(site_rain_data$Site_Name)

```

Now expand to a daily dataFrame

We'll:

```         
Convert each sample into a sequence of dates based on the collection period

Distribute rainfall equally across days in the period (excluding the final collection day)

Assign the same isotope value (d18O, d2H) to each day in the period

Then we'll combine all samples into one daily-level dataframe
```

```{r}


# Ensure dates are properly formatted (you already have this from glimpse)
daily_expanded <- site_rain_data %>%
  # Calculate the number of days in each sampling period (excluding collection day)
  mutate(
    Start_Date = as.Date(Start_Date),
    Collection_Date = as.Date(Collection_Date),
    days = as.numeric(Collection_Date - Start_Date)
  ) %>%
  # Filter out any zero-day or malformed samples
  filter(days > 0 & !is.na(Precipitation_mm)) %>%
  # Expand each row into daily records
  rowwise() %>%
  mutate(
    date_seq = list(seq.Date(Start_Date, Collection_Date - 1, by = "day")),
    rain_per_day = Precipitation_mm / days
  ) %>%
  ungroup() %>%
  unnest(cols = c(date_seq)) %>%
  mutate(
    Precipitation_mm = rain_per_day,
    d18O = as.numeric(d18O),
    d2H = as.numeric(d2H)
  ) %>%
  select(Site_Name, Project, date = date_seq, Precipitation_mm, d18O, d2H)

```

Rainfall-Weighted Monthly Means (calendar months)

Now we'll:

```         
Extract year/month

Group by site and calendar month

Calculate rainfall-weighted means for d18O and d2H
```

```{r}
rainfall_weighted_monthly <- daily_expanded %>%
  mutate(year_month = floor_date(date, unit = "month")) %>%
  group_by(Site_Name, Project, year_month) %>%
  summarise(
    total_rain = sum(Precipitation_mm, na.rm = TRUE),
    
    weighted_d18O = if (all(is.na(d18O) | Precipitation_mm == 0)) NA_real_ else
      sum(d18O * Precipitation_mm, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O)], na.rm = TRUE),
    
    weighted_d2H = if (all(is.na(d2H) | Precipitation_mm == 0)) NA_real_ else
      sum(d2H * Precipitation_mm, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H)], na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  # If total rainfall is zero, set isotope values to NA
  mutate(
    weighted_d18O = ifelse(total_rain == 0, NA_real_, weighted_d18O),
    weighted_d2H  = ifelse(total_rain == 0, NA_real_, weighted_d2H)
  )


```

Now - Combine monthly_weighted_flagged with rainfall_weighted_monthly and then adjust year_month to refer to the previous month

```{r}

# Step 1: Rename VCSN-derived columns
monthly_flagged_renamed <- monthly_weighted_flagged %>%
  rename(
    VCSN_total_rainfall = total_rainfall,
    VCSN_weighted_d2H = weighted_d2H,
    VCSN_weighted_d18O = weighted_d18O
  )


# Perform a full join to bring the adjusted rainfall and isotope values into the VCSN dataframe
monthly_combined <- full_join(
  monthly_flagged_renamed,
  rainfall_weighted_monthly %>%
    select(Site_Name, year_month, 
           Adjusted_total_rain = total_rain, 
           Adjusted_weighted_d18O = weighted_d18O, 
           Adjusted_weighted_d2H = weighted_d2H),
  by = c("Site_Name", "year_month")
)


```

Now cut the dataframe down for plotting

```{r}
filtered_df <- monthly_combined %>%
  filter(!is.na(Adjusted_total_rain))

```

### Check to see why the plots below are missing a couple of sites in the html file...

```{r}
unique(filtered_df$Site_Name)
```

## Comparison plots and RMSE calculations

```{r weighting plots through time, warning=FALSE}


# Calculate RMSE values for each variable
rmse_rain <- sqrt(mean((filtered_df$VCSN_total_rainfall - filtered_df$Adjusted_total_rain)^2, na.rm = TRUE))
rmse_d18O <- sqrt(mean((filtered_df$VCSN_weighted_d18O - filtered_df$Adjusted_weighted_d18O)^2, na.rm = TRUE))
rmse_d2H  <- sqrt(mean((filtered_df$VCSN_weighted_d2H - filtered_df$Adjusted_weighted_d2H)^2, na.rm = TRUE))

# Plot for Rainfall
rain_plot <- ggplot(filtered_df, aes(x = Adjusted_total_rain, y = VCSN_total_rainfall, color = Site_Name)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = paste0("Monthly Rainfall: RMSE = ", round(rmse_rain, 2), " mm"),
       x = "Measured Rainfall (mm)",
       y = "VCSN Rainfall (mm)") +
  theme_minimal()

# Plot for d18O
d18O_plot <- ggplot(filtered_df, aes(x = Adjusted_weighted_d18O, y = VCSN_weighted_d18O, color = Site_Name)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = paste0("Monthly d18O: RMSE = ", round(rmse_d18O, 2), " ‰"),
       x = "Measured d18O (‰)",
       y = "VCSN d18O (‰)") +
  theme_minimal()

# Plot for d2H
d2H_plot <- ggplot(filtered_df, aes(x = Adjusted_weighted_d2H, y = VCSN_weighted_d2H, color = Site_Name)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = paste0("Monthly d2H: RMSE = ", round(rmse_d2H, 2), " ‰"),
       x = "Measured d2H (‰)",
       y = "VCSN d2H (‰)") +
  theme_minimal()

# Print plots
print(rain_plot)
print(d18O_plot)
print(d2H_plot)


```

### Line plots

```{r}
# --- Combine d2H and d18O into one long dataframe ---

# Fix the date frameshift

filtered_df <- filtered_df %>%
  mutate(year_month = year_month %m+% months(1))

# and plot

df_combined <- filtered_df %>%
  select(Site_Name, year_month,
         VCSN_weighted_d2H, Adjusted_weighted_d2H,
         VCSN_weighted_d18O, Adjusted_weighted_d18O) %>%
  pivot_longer(
    cols = c(VCSN_weighted_d2H, Adjusted_weighted_d2H,
             VCSN_weighted_d18O, Adjusted_weighted_d18O),
    names_to = c("Source", "isotope"),
    names_pattern = "(VCSN|Adjusted)_weighted_(d2H|d18O)",
    values_to = "value"
  ) %>%
  mutate(
    isotope = recode(isotope,
                     "d2H" = "δ²H",
                     "d18O" = "δ¹⁸O"),
    Source = recode(Source,
                    "VCSN" = "VCSN weighted",
                    "Adjusted" = "Adjusted weighted")
  )

# Reorder Site_Name so MAIMAI (M8) comes last
df_combined <- df_combined %>%
  mutate(Site_Name = factor(Site_Name,
                            levels = c("Ashley Dene",
                                       "B3. Kaitaia",
                                       "B3.Christchurch",
                                       "Melrose Farm",
                                       "MAIMAI (M8)")))

ggplot(df_combined, aes(x = year_month, y = value, color = Source)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  facet_grid(
    isotope ~ Site_Name,
    scales = "free",
    space = "fixed"
  ) +
  labs(
    x = "Date",
    # Multiline isotope label for both rows, spaced nicely
    y = expression(paste("δ"^2*"H (‰)                    ", "\n\n\n", "δ"^18*"O (‰)")),
    color = "Data Source"
  ) +
  scale_x_date(
    date_labels = "%b-%y",
    breaks = scales::breaks_pretty(n = 5),
    expand = expansion(mult = c(0.01, 0.05))
  ) +
  scale_color_manual(
    name = "Rainfall amount data source",
    values = c(
      "VCSN weighted" = "#0072B2",
      "Adjusted weighted" = "#D55E00"
    ),
    labels = c(
      "VCSN weighted" = "VCSN rainfall",
      "Adjusted weighted" = "On-site rainfall"
    )
  ) +  # ✅ removed the stray parenthesis before this plus sign
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    strip.text = element_text(size = 10),
    strip.text.y = element_blank(),  # remove isotope labels on side
    legend.position = "bottom",
    panel.spacing = unit(1.2, "lines"),
    plot.title = element_text(hjust = 0.5)
  )

# --- Save at publication quality (600 dpi) ---
ggsave(
  filename = "isotope_comparison_combined VCSN.tif",
  plot = last_plot(),
  width = 10,
  height = 6,
  dpi = 600,
  units = "in",
  compression = "lzw"
)


```

# Test for effects of collection date 'frame shifts' away from a full month sampling date

```{r }
library(dplyr)
library(lubridate)
library(tidyr)

# Step 1: Calendar-month weighted isotope values (true)
monthly_true <- daily_expanded %>%
  filter(!is.na(d2H) | !is.na(d18O)) %>%
  mutate(
    data_month = floor_date(date, "month"),
    Collection_date = ceiling_date(date, "month")
  ) %>%
  group_by(Site_Name, data_month, Collection_date) %>%
  summarise(
    weighted_d2H = if (all(is.na(d2H))) NA else sum(Precipitation_mm * d2H, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H)], na.rm = TRUE),
    weighted_d18O = if (all(is.na(d18O))) NA else sum(Precipitation_mm * d18O, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O)], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year_month = Collection_date,
    Start_date = data_month
  ) %>%
  select(Site_Name, year_month, weighted_d2H, weighted_d18O)

# Step 2: Add one month before and after per site
monthly_padded <- monthly_true %>%
  group_by(Site_Name) %>%
  summarise(
    start = min(year_month) %m-% months(1),
    end = max(year_month) %m+% months(1),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    full_months = list(seq.Date(start, end, by = "1 month"))
  ) %>%
  unnest(cols = c(full_months)) %>%
  rename(year_month = full_months) %>%
  mutate(
    Collection_date = year_month,
    Start_date = year_month %m-% months(1)
  ) %>%
  select(Site_Name, year_month, Start_date, Collection_date) %>%
  left_join(monthly_true, by = c("Site_Name", "year_month")) %>%
  arrange(Site_Name, year_month)

## Add in frame shift columns

monthly_padded <- monthly_padded %>%
  mutate(
    Start_date5 = Start_date + days(5),
    Collection_date5 = Collection_date + days(5),
    Start_date10 = Start_date + days(10),
    Collection_date10 = Collection_date + days(10),
    Start_date15 = Start_date + days(15),
    Collection_date15 = Collection_date + days(15)
  )




```

### Join new date ranges back to daily dataset and calculate frame-shifted values

```{r}
library(dplyr)
library(purrr)

site_list <- unique(monthly_padded$Site_Name)

# Function to compute shifted values for one site
compute_shifts <- function(site_name) {
  df_site <- monthly_padded %>% filter(Site_Name == site_name)

  df_site <- df_site %>%
    rowwise() %>%
    mutate(
      d2H_shift5 = {
        df <- daily_expanded %>%
          filter(
            Site_Name == site_name,
            date >= Start_date5,
            date <= Collection_date5,
            !is.na(d2H)
          )
        if (nrow(df) == 0) NA_real_ else sum(df$Precipitation_mm * df$d2H) / sum(df$Precipitation_mm)
      },
      d18O_shift5 = {
        df <- daily_expanded %>%
          filter(
            Site_Name == site_name,
            date >= Start_date5,
            date <= Collection_date5,
            !is.na(d18O)
          )
        if (nrow(df) == 0) NA_real_ else sum(df$Precipitation_mm * df$d18O) / sum(df$Precipitation_mm)
      }
    ) %>%
    ungroup()

  return(df_site)
}

# Apply the function to all sites
monthly_padded <- map_dfr(site_list, compute_shifts)

monthly_padded$d2H_shift10 <- NA_real_
monthly_padded$d18O_shift10 <- NA_real_

for (site in unique(monthly_padded$Site_Name)) {
  site_rows <- monthly_padded$Site_Name == site
  for (i in which(site_rows)) {
    start <- monthly_padded$Start_date10[i]
    end <- monthly_padded$Collection_date10[i]

    df <- daily_expanded %>%
      filter(Site_Name == site, date >= start, date <= end, !is.na(d2H))

    if (nrow(df) > 0) {
      monthly_padded$d2H_shift10[i] <- sum(df$Precipitation_mm * df$d2H) / sum(df$Precipitation_mm)
    }

    df <- daily_expanded %>%
      filter(Site_Name == site, date >= start, date <= end, !is.na(d18O))

    if (nrow(df) > 0) {
      monthly_padded$d18O_shift10[i] <- sum(df$Precipitation_mm * df$d18O) / sum(df$Precipitation_mm)
    }
  }
}

monthly_padded$d2H_shift15 <- NA_real_
monthly_padded$d18O_shift15 <- NA_real_

for (site in unique(monthly_padded$Site_Name)) {
  site_rows <- monthly_padded$Site_Name == site
  for (i in which(site_rows)) {
    start <- monthly_padded$Start_date15[i]
    end <- monthly_padded$Collection_date15[i]

    df <- daily_expanded %>%
      filter(Site_Name == site, date >= start, date <= end, !is.na(d2H))

    if (nrow(df) > 0) {
      monthly_padded$d2H_shift15[i] <- sum(df$Precipitation_mm * df$d2H) / sum(df$Precipitation_mm)
    }

    df <- daily_expanded %>%
      filter(Site_Name == site, date >= start, date <= end, !is.na(d18O))

    if (nrow(df) > 0) {
      monthly_padded$d18O_shift15[i] <- sum(df$Precipitation_mm * df$d18O) / sum(df$Precipitation_mm)
    }
  }
}


```

### Now make a daily dataframe from the monthly weighted data

```{r}

library(dplyr)
library(tidyr)
daily_weighted <- monthly_padded %>%
  rowwise() %>%
  mutate(date_seq = list(seq.Date(Start_date, Collection_date - 1, by = "day"))) %>%
  unnest(date_seq) %>%
  dplyr::select(Site_Name, Date = date_seq, weighted_d2H, weighted_d18O)

# 5-day shifted
daily_weighted_5 <- monthly_padded %>%
  rowwise() %>%
  mutate(date_seq = list(seq.Date(Start_date5, Collection_date5 - 1, by = "day"))) %>%
  unnest(date_seq) %>%
  dplyr::select(Site_Name, Date = date_seq, d2H = d2H_shift5, d18O = d18O_shift5)

# 10-day shifted
daily_weighted_10 <- monthly_padded %>%
  rowwise() %>%
  mutate(date_seq = list(seq.Date(Start_date10, Collection_date10 - 1, by = "day"))) %>%
  unnest(date_seq) %>%
  dplyr::select(Site_Name, Date = date_seq, d2H = d2H_shift10, d18O = d18O_shift10)

# 15-day shifted
daily_weighted_15 <- monthly_padded %>%
  rowwise() %>%
  mutate(date_seq = list(seq.Date(Start_date15, Collection_date15 - 1, by = "day"))) %>%
  unnest(date_seq) %>%
  dplyr::select(Site_Name, Date = date_seq, d2H = d2H_shift15, d18O = d18O_shift15)


library(dplyr)

# First, rename the isotope columns in the shifted datasets
daily_weighted_5_renamed <- daily_weighted_5 %>%
  rename(d2H_5 = d2H,
         d18O_5 = d18O)

daily_weighted_10_renamed <- daily_weighted_10 %>%
  rename(d2H_10 = d2H,
         d18O_10 = d18O)

daily_weighted_15_renamed <- daily_weighted_15 %>%
  rename(d2H_15 = d2H,
         d18O_15 = d18O)

# Start with daily_weighted, then left_join the others by Site_Name and Date
combined_daily <- daily_weighted %>%
  left_join(daily_weighted_5_renamed,  by = c("Site_Name", "Date")) %>%
  left_join(daily_weighted_10_renamed, by = c("Site_Name", "Date")) %>%
  left_join(daily_weighted_15_renamed, by = c("Site_Name", "Date"))

# View result
glimpse(combined_daily)



```

Now reattach this to the original daily dataframe

```{r}
library(dplyr)

# Make sure the date columns are of the same name and type
combined_daily_renamed <- combined_daily %>%
  rename(date = Date)

# Now join the new data to daily_expanded
daily_expanded_joined <- daily_expanded %>%
  left_join(combined_daily_renamed %>%
              select(Site_Name, date, weighted_d2H, weighted_d18O, 
                     d2H_5, d18O_5, d2H_10, d18O_10, d2H_15, d18O_15),
            by = c("Site_Name", "date"))

```

### Put together a dataframe of 'frame shifted' dates

```{r}

monthly_final <- daily_expanded_joined %>%
  # keep any row that has at least one isotope value
  filter(
    !is.na(d2H) | !is.na(d18O) |
    !is.na(d2H_5) | !is.na(d18O_5) |
    !is.na(d2H_10) | !is.na(d18O_10) |
    !is.na(d2H_15) | !is.na(d18O_15)
  ) %>%
  mutate(
    data_month     = floor_date(date,  "month"),
    Collection_date = ceiling_date(date, "month")
  ) %>%
  group_by(Site_Name, data_month, Collection_date) %>%
  summarise(
    # — True daily isotopes
    weighted_d2H        = if (all(is.na(d2H)))       NA_real_ else sum(Precipitation_mm * d2H,    na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H)],    na.rm = TRUE),
    weighted_d18O       = if (all(is.na(d18O)))      NA_real_ else sum(Precipitation_mm * d18O,   na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O)],   na.rm = TRUE),
    # — Check (using already-weighted daily values)
    weighted_d2H_check  = if (all(is.na(weighted_d2H)))  NA_real_ else sum(Precipitation_mm * weighted_d2H,  na.rm = TRUE) / sum(Precipitation_mm[!is.na(weighted_d2H)],  na.rm = TRUE),
    weighted_d18O_check = if (all(is.na(weighted_d18O))) NA_real_ else sum(Precipitation_mm * weighted_d18O, na.rm = TRUE) / sum(Precipitation_mm[!is.na(weighted_d18O)], na.rm = TRUE),

    # — 5‑day shifted
    weighted_d2H_5      = if (all(is.na(d2H_5)))     NA_real_ else sum(Precipitation_mm * d2H_5,   na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H_5)],   na.rm = TRUE),
    weighted_d18O_5     = if (all(is.na(d18O_5)))    NA_real_ else sum(Precipitation_mm * d18O_5,  na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O_5)],  na.rm = TRUE),

    # — 10‑day shifted
    weighted_d2H_10     = if (all(is.na(d2H_10)))    NA_real_ else sum(Precipitation_mm * d2H_10,  na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H_10)],  na.rm = TRUE),
    weighted_d18O_10    = if (all(is.na(d18O_10)))   NA_real_ else sum(Precipitation_mm * d18O_10, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O_10)], na.rm = TRUE),

    # — 15‑day shifted
    weighted_d2H_15     = if (all(is.na(d2H_15)))    NA_real_ else sum(Precipitation_mm * d2H_15,  na.rm = TRUE) / sum(Precipitation_mm[!is.na(d2H_15)],  na.rm = TRUE),
    weighted_d18O_15    = if (all(is.na(d18O_15)))   NA_real_ else sum(Precipitation_mm * d18O_15, na.rm = TRUE) / sum(Precipitation_mm[!is.na(d18O_15)], na.rm = TRUE),

    .groups = "drop"
  ) %>%
  # finally label your "month" and "start"
  mutate(
    year_month = Collection_date,
    Start_date = data_month
  ) %>%
  select(
    Site_Name, year_month, Start_date,
    weighted_d2H,    weighted_d18O,
    weighted_d2H_check,    weighted_d18O_check,
    weighted_d2H_5,  weighted_d18O_5,
    weighted_d2H_10, weighted_d18O_10,
    weighted_d2H_15, weighted_d18O_15
  )



```

### Plotting and RMSE calcs for date-shifting effects

### Reshape monthly_final for plotting

This reformats the wide structure into a tidy long format with labeled shift windows:

```{r}

comparison_long <- monthly_final %>%
  select(Site_Name, year_month, starts_with("weighted_d")) %>%
  pivot_longer(
    cols = starts_with("weighted_d"),
    names_to = "parameter",
    values_to = "value"
  ) %>%
  mutate(
    isotope = case_when(
      grepl("d2H", parameter) ~ "δ²H",
      grepl("d18O", parameter) ~ "δ¹⁸O"
    ),
    type = case_when(
      parameter %in% c("weighted_d2H", "weighted_d18O") ~ "Calendar month",
      parameter %in% c("weighted_d2H_5", "weighted_d18O_5") ~ "5-day shift",
      parameter %in% c("weighted_d2H_10", "weighted_d18O_10") ~ "10-day shift",
      parameter %in% c("weighted_d2H_15", "weighted_d18O_15") ~ "15-day shift",
      grepl("check", parameter) ~ NA_character_,
      TRUE ~ NA_character_
    ),
    type = factor(type, levels = c("Calendar month", "5-day shift", "10-day shift", "15-day shift"))
  ) %>%
  filter(!is.na(type)) %>%
  mutate(Site_Name = factor(Site_Name, levels = c(setdiff(unique(Site_Name), "MAIMAI (M8)"), "MAIMAI (M8)"))) %>%
  filter(!(Site_Name == "MAIMAI (M8)" & isotope == "δ¹⁸O"))

```

### Plot the isotope values by shift type

```{r}

# Keep isotope labels clean (no units here — we’ll add them manually in y-axis label)
comparison_long <- comparison_long %>%
  mutate(isotope = recode(isotope,
                          "δ²H"  = "δ²H",
                          "δ¹⁸O" = "δ¹⁸O"))

ggplot(comparison_long, aes(x = year_month, y = value, color = type)) +
  geom_line(size = 1) +
  facet_grid(
    isotope ~ Site_Name,
    scales = "free",
    space = "fixed"
    # removed switch = "y" so isotope names disappear from the side
  ) +
  labs(
    x = "Date",
    # Custom multiline isotope label (adjust spacing as needed)
    y = expression(paste("δ"^2*"H (‰)                    ", "\n\n\n", "δ"^18*"O (‰)")),
    color = "Sampling period"
  ) +
  scale_x_date(
    date_labels = "%b-%y",
    breaks = scales::breaks_pretty(n = 5),
    expand = expansion(mult = c(0.01, 0.05))
  ) +
  scale_color_manual(
    values = c(
      "Calendar month" = "#0072B2",
      "5-day shift"    = "#D55E00",
      "10-day shift"   = "#009E73",
      "15-day shift"   = "#CC79A7"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    strip.text = element_text(size = 10),
    strip.text.y = element_blank(),  # removes the facet row labels
    legend.position = "bottom",
    panel.spacing = unit(1.2, "lines"),
    plot.title = element_text(hjust = 0.5)
  )

# Save high-resolution version
ggsave(
  filename = "isotope_comparison.tif",
  plot = last_plot(),
  width = 10,
  height = 6,
  dpi = 600,
  units = "in",
  compression = "lzw"
)

```

### Calculate RMSE values

This compares each shifted series against the calendar-month baseline for both δ²H and δ¹⁸O

```{r}
library(Metrics)
library(dplyr)
library(tidyr)

rmse_results <- comparison_long %>%
  pivot_wider(
    id_cols = c(Site_Name, year_month, isotope),
    names_from = type,
    values_from = value
  ) %>%
  group_by(Site_Name, isotope) %>%
  summarise(
    RMSE_5  = rmse(`Calendar month`, `5-day shift`),
    RMSE_10 = rmse(`Calendar month`, `10-day shift`),
    RMSE_15 = rmse(`Calendar month`, `15-day shift`),
    .groups = "drop"
  )

# Add overall RMSE (pooled)
overall_rmse <- comparison_long %>%
  pivot_wider(
    id_cols = c(Site_Name, year_month, isotope),
    names_from = type,
    values_from = value
  ) %>%
  group_by(isotope) %>%
  summarise(
    Site_Name = "Overall (pooled)",
    RMSE_5  = rmse(`Calendar month`, `5-day shift`),
    RMSE_10 = rmse(`Calendar month`, `10-day shift`),
    RMSE_15 = rmse(`Calendar month`, `15-day shift`),
    .groups = "drop"
  )

# Optional: overall mean of site-level RMSEs
mean_rmse <- rmse_results %>%
  group_by(isotope) %>%
  summarise(
    Site_Name = "Mean of site RMSEs",
    RMSE_5  = mean(RMSE_5, na.rm = TRUE),
    RMSE_10 = mean(RMSE_10, na.rm = TRUE),
    RMSE_15 = mean(RMSE_15, na.rm = TRUE),
    .groups = "drop"
  )

# Combine
rmse_results_combined <- bind_rows(rmse_results, mean_rmse, overall_rmse)

print(rmse_results_combined)




```

# Combined RMSE table including rainfall weighting

```{r}
# ---- NEW: Calculate VCSN weighting RMSEs ----
# ---- Helper: safe RMSE function that removes NAs ----
safe_rmse <- function(x, y) {
  df <- na.omit(data.frame(x, y))
  if (nrow(df) == 0) return(NA_real_)
  rmse(df$x, df$y)
}

# ---- NEW: Calculate VCSN weighting RMSEs ----
vcsn_rmse <- df_combined %>%
  pivot_wider(
    id_cols = c(Site_Name, year_month, isotope),
    names_from = Source,
    values_from = value
  ) %>%
  group_by(Site_Name, isotope) %>%
  summarise(
    RMSE_VCSN = safe_rmse(`Adjusted weighted`, `VCSN weighted`),
    .groups = "drop"
  )

vcsn_mean <- vcsn_rmse %>%
  group_by(isotope) %>%
  summarise(
    Site_Name = "Mean of site RMSEs",
    RMSE_VCSN = mean(RMSE_VCSN, na.rm = TRUE),
    .groups = "drop"
  )

vcsn_overall <- df_combined %>%
  pivot_wider(
    id_cols = c(Site_Name, year_month, isotope),
    names_from = Source,
    values_from = value
  ) %>%
  group_by(isotope) %>%
  summarise(
    Site_Name = "Overall (pooled)",
    RMSE_VCSN = safe_rmse(`Adjusted weighted`, `VCSN weighted`),
    .groups = "drop"
  )

vcsn_rmse_combined <- bind_rows(vcsn_rmse, vcsn_mean, vcsn_overall)


# ---- Merge with main RMSE table ----
rmse_results_final <- rmse_results_combined %>%
  left_join(vcsn_rmse_combined, by = c("Site_Name", "isotope"))

# ---- Round results for presentation ----
rmse_results_final <- rmse_results_final %>%
  mutate(across(starts_with("RMSE"), ~round(.x, 2)))

print(rmse_results_final)
```

# Write final 'Stage 1' data outputs to an excel file

```{r}
library(writexl)

# Function to clean illegal XML characters
clean_illegal_chars <- function(df) {
  df[] <- lapply(df, function(x) {
    if (is.character(x)) {
      # Force encoding to UTF-8 and replace invalid characters
      x <- iconv(x, from = "", to = "UTF-8", sub = "")
      # Remove characters not allowed in XML 1.0
      x <- gsub("[^\x09\x0A\x0D\x20-\x7F]", "", x)
    }
    return(x)
  })
  return(df)
}


# Clean each dataframe
Lookup_filtered_clean <- clean_illegal_chars(Lookup_filtered)
Full_database_clean <- clean_illegal_chars(Full_database)
monthly_weighted_clean <- clean_illegal_chars(monthly_weighted)
summary_by_project_clean <- clean_illegal_chars(summary_by_project)

# --- Prepare for export with consistent names ---

# Fix column names in the monthly dataframe
monthly_weighted_clean_fixed <- monthly_weighted_clean %>%
  rename(
    Site = Site_Name
  ) %>%
  mutate(day = 1) %>%  # add first of month as placeholder day
  relocate(day, .after = month)

# Fix the lookup dataframe name and columns to match "SITE DATA lookup"
Lookup_filtered_clean_fixed <- Lookup_filtered_clean %>%
  rename(
    Site = Site_Name,
    lat = lat,
    long = long,
    Elevation_masl = Elevation_masl
  )

# --- Write to Excel with names matching your import script ---
write_xlsx(
  list(
    "SITE DATA lookup" = Lookup_filtered_clean_fixed,
    "Monthly" = monthly_weighted_clean_fixed,
    "Subdaily data" = ddata,
    "Daily and greater" = Full_database_clean,
    "Datatypes" = summary_by_project_clean
  ),
  path = "output/Processesed_database.xlsx"
)






```

# Clean up metadata table for publishing

```{r}
# Load partially cleaned metadata table

loaded.metadata<-read.csv("metadata_table.csv")

# Elevation extraction using elevatr

library(elevatr)
library(sf)
library(dplyr)


# Convert to sf
loaded.metadata.sf <- st_as_sf(
  loaded.metadata,
  coords = c("long", "lat"),
  crs = 4326,
  remove = FALSE
)

elevations <- get_elev_point(
  locations = loaded.metadata.sf,
  src = "aws",
  z = 6             # coarse (~1 km) but will download safely
)

# Add elevation (meters)
loaded.metadata$Elev_external <- elevations$elevation



```

## Add earliest dates and most recent dates onto lookup table

```{r}
library(dplyr)

# ---- 1. Summarise isotope data by Site_Name ----

site_summary <- Full_database %>%
  group_by(Site_Name) %>%
  summarise(
    Date_first_collection = min(Collection_Date, na.rm = TRUE),
    Date_last_collection  = max(Collection_Date, na.rm = TRUE),

    # Count rows with data (non-NA)
    n_d18O_precip_measurement = sum(!is.na(d18O)),
    n_d2H_precip_measurement  = sum(!is.na(d2H)),
    .groups = "drop"
  )

# ---- 2. Join to metadata ----

loaded.metadata <- loaded.metadata %>%
  left_join(site_summary, by = "Site_Name")

```

## Add in abbreviated project name for Figure 1

```{r}
metadata2<-read.csv("input/Metadata readin.csv")
metadata2 <- metadata2 |>
  mutate(Project = stringr::str_replace_all(Project, "\x96", "–"))

project_name_map <- c(
  "CDRP" = "CDRP",
  "None" = "TUD",
  "GNIP/M/NZ/01" = "GNIP",
  "Forest Flows" = "Forest Flows",
  "Mt TeAroha " = "Mt Te Aroha",
  "Waihi Estuary Focus Catchment Investigation" = "Waihi Estuary",
  "Land Use Intensification: Sustainable Management of Water Quantity and Quality Programme (NIWA)" = "LUWQ NIWA",
  "Land Use Intensi?cation: Sustainable Management of Water Quantity and Quality Programme (NIWA)" = "LUWQ NIWA",  # old spelling
  "Rotorua Lakes FaCTS" = "Rotorua FaCTS",
  "NZWaM" = "NZWaM",
  "Taylor 1990" = "DSIR",
  "Tritium Input Function" = "GNS Tritium Input Function",
  "Phase1_early M8 catchment-scale research and observations (1974–1988)" = "Maimai M8",
  "Hill AP transect" = "Hill AP transect",
  "Tim Kerr postdoc" = "Kerr Postdoc",
  "Hill AP transect / Jacks Thesis" = "Hill AP transect",
  "Waterscape" = "Waterscape",
  "Jack thesis + Zane thesis" = "UC Postgrad",
  "AOSnow" = "Hill AOSnow",
  "Jack thesis" = "UC Postgrad",
  "Ashley Dene" = "Ashley Dene",
  "Hill Haast transect" = "Hill Haast transect",
  "Environment Southland" = "Environment Southland",
  "B3" = "B3",
  "Nehemy/Simmons tree water sources " = "Trent University"
)

# Add the new column to metadata2
metadata2 <- metadata2 |>
  dplyr::mutate(
    `Figure 1 project abbreviation` = dplyr::recode(Project, !!!project_name_map)
  )


```

### Identify project names that don't match

```{r}
setdiff(metadata2$Project, names(project_name_map))

head(metadata2$`Figure 1 project abbreviation`)

metadata2 <- metadata2 |>
  relocate('Figure 1 project abbreviation', .after = Project)

write.csv(metadata2, "metadata2.csv", row.names = FALSE)
```

### Pull out metadata information for manuscript text

```{r}
# Identify snowpack sites
metadata2 <- metadata2 %>%
  mutate(is_snowpack = str_detect(tolower(Notes), "snowpack"))

total_sites <- nrow(metadata2)
snow_sites  <- sum(metadata2$is_snowpack, na.rm = TRUE)
non_snow_sites <- total_sites - snow_sites

total_sites
snow_sites
non_snow_sites

```

Earliest and latest collection dates

```{r}
library(lubridate)

metadata2 <- metadata2 %>%
  mutate(
    Date_first_collection = dmy(Date_first_collection),
    Date_last_collection  = dmy(Date_last_collection)
  )

earliest_first <- min(metadata2$Date_first_collection, na.rm = TRUE)
latest_last    <- max(metadata2$Date_last_collection,  na.rm = TRUE)

earliest_first
latest_last

```

Max and average number of samples at non 'snowpack' sites

```{r}
# Maximum samples at any site
max_samples <- max(metadata2$n_d18O_precip_measurement, na.rm = TRUE)

# Average samples per *non-snowpack* site
avg_samples_non_snow <- metadata2 %>%
  filter(!is_snowpack) %>%
  summarise(mean_samples = median(n_d18O_precip_measurement, na.rm = TRUE)) %>%
  pull(mean_samples)

max_samples
avg_samples_non_snow

```

### Frequency breakdown

```{r}

freq_split_non_snow <- metadata2 %>%
  filter(!is_snowpack) %>%        # keep only non-snowpack sites
  count(Frequency, name = "n_sites")

freq_split_non_snow

```
